Variation generation and mean fitness increase: \\ Necessity is the mother of genetic invention

### 2017-11-16 ms.tex
primitive condition; I typeface; nameref emph

% Yoav & Lee commands
\newcommand*{\tr}{^\intercal}
\let\vec\mathbf
\newcommand{\matrx}[1]{{\left[ \stackrel{}{#1}\right]}}
\newcommand{\diag}[1]{\mbox{diag}\matrx{#1}}
\newcommand{\goesto}{\rightarrow}
\newcommand{\dspfrac}[2]{\frac{\displaystyle #1}{\displaystyle #2} }
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}
\newtheorem{lemma}{Lemma}
\newtheorem{remark}{Remark}
%\newtheorem*{mfp}{Theorem: Mean Fitness Principle}
%\newtheorem*{rvp}{Corollary: Reproductive Value Principle}
%\newtheorem*{frv}{Fisher's reproductive values}
\renewcommand\qedsymbol{} % no square at end of proof
\newcommand{\cl}{\mathbf{L}}
\newcommand{\cj}{\mathbf{J}}
\newcommand{\ci}{\mathbf{I}}

@@

\paragraph*{General model.}\label{sec:general_model}
Consider a large haploid population with an arbitrary set of types $A_1, A_2, \ldots, A_n$.
The frequency and fitness of individuals of type $A_k$ are $f_k$ and $w_k$, respectively.
The probability that an individual of type $A_k$ will transition to some other type is $C_k$, and given a transition occurs, the probability that it will transition to type $A_j$ is $M_{j,k}$.
Therefore, the change in the frequencies of type $A_k$ is described by the transformation $f \to f'$: 
\begin{equation}
\bar{w} f'_k = (1-C_k) w_k f_k + \sum_{j=1}^{n}{C_j M_{k,j} w_j f_j},
\label{eq:model_sum}
\end{equation}
or in matrix form
\begin{equation}
\bar{w} f' = (\ci - C + MC)D f,
\label{eq:model}
\end{equation}
where $f=(f_1, f_2, \ldots, f_n)$ is a stochastic vector such that $\sum_{k=1}^n{f_k} = 1$;
$D$ is a diagonal matrix with diagonal elements $w_k$ such that $w_k \ne w_j$ for $k \ne j$, and we can assume $0 < w_k \le 1$, i.e., fitness values are relative to the fitness of the fittest type; 
$C$ is a diagonal non-negative matrix with diagonal elements $C_k$;
$M$ is a column-stochastic matrix: $M_{j,k} \ge 0$ for all $j,k$ and $\sum_{j=1}^n {M_{j,k}} = 1$ for all $k$; 
$\ci$ is the identity matrix;
and $\bar{w}$ is the normalizing factor such that $\sum_{k=1}^n{f'_k}=1$ and is equal to the population mean fitness (see~\nameref{sec:AppA}).

@@

The initial phenotype is acquired by learning the mature phenotype of one of the parents (i.e. uni-parental vertical transmission).
In individuals with initial phenotype $k$, the mature phenotype is the same as the initial phenotype with probability $1-C_k$, and is modified by individual learning or \emph{exploration}~\citep{Borenstein2008} with probability $C_k$.
Such individual exploratory learning, which can be considered intentional or the result of transmission infidelity, modifies initial phenotype $k$ to mature phenotype $j$ with probability $M_{k,j}$.

For example, the case of \emph{symmetric individual learning}~\citep{Borenstein2008} can be represented by
\begin{equation}
M_{k,j} = \begin{cases}
\frac{1}{2b+1}, & \abs{k-j} \le b \\
0, & otherwise
\end{cases},
\end{equation}
where the breadth of exploration is $b \ge 0$, and phenotypes $0$ and $n$ "absorb" transitions to $k<0$ and $k>n$, respectively. 
For instance, for $n=5$ and $b=1$ we have
\begin{equation}
M = \begin{bmatrix}
2/3 & 1/3 & 0 & 0 & 0 \\
1/3 & 1/3 & 1/3 & 0 & 0 \\
0 & 1/3 & 1/3 & 1/3 & 0 \\
0 & 0 & 1/3 & 1/3 & 1/3 \\
0 & 0 & 0 & 1/3 & 2/3
\end{bmatrix}.
\end{equation}

Therefore, if the frequency of individuals with mature phenotype $k$ in the current generation is $f_k$, then the frequency in the next generation $f'_k$ is
\begin{equation}
f'_k = (1-C_k) f_k \frac{w_{\abs{k-h}}}{\bar{w}} \sum_{j=1}^{n}{C_j M_{k,j} \frac{w_{\abs{j-h}}}{\bar{w}} f_j },
\end{equation}
where $\bar{w} = \sum_{k=1}^{n}{f_k w_{\abs{k-h}}}$ is the population mean fitness.

@@

\begin{lemma}[Mean fitness principle]\label{lemma:mfp}
Let $\bar{w}^*$ be the leading eigenvalue of $(I-C+MC)D$,
and $u$ and $v$ be the corresponding left and right eigenvectors.
%TODO If $MCD$ is primitive, then
Then,
\begin{equation}\label{mfp}
\frac{\partial \bar{w}^*}{\partial C_k} = 
\frac{u_k v_k}{C_k} (\bar{w}^* - w_k),
\end{equation}
or in simpler terms,
\begin{equation}\label{mfp_sign}
sign\frac{\partial \bar{w}^*}{\partial C_k} = 
sign(\bar{w}^* - w_k).
\end{equation}

@@

% see Altenberg 2012 Ecological Monographs for a condition-dependent migration model
We have given several examples of variation-generating processes for which this principle applies -- namely mutation, migration, and learning, see ~\nameref{sec:models} -- but our model may apply to other processes as well.
For example, the \emph{reduction principle} applies to ecological models of dispersal, and~\citet{Gueijman2013} have demonstrated that even in homogeneous environments, fitness-associated dispersal increases the mean fitness of diploid populations and is favored by selection over uniform dispersal.
Similarly, if the transmission fidelity of culturally-transmitted traits depends on the type or fitness of the transmitting individual, we expect that our results will hold (see~\nameref{sec:learn_model}).

Eq.~\ref{mfp} is a generalization of a result of~\citet[Eq.~4]{Ram2012}.
\citeauthor{Ram2012} modeled the accumulation of mutant alleles in a population evolving at the mutation-selection balance (see \nameref{sec:mut_model_2}).
Using Eq.~\ref{eq:Caswells_formula} in \nameref{sec:AppB} and a recursion on the ratios of the reproductive values \cite[see][eqs.~A5-6]{Ram2012}, they concluded that at the mutation-selection balance, if individuals with below-average fitness ($D_k < \bar{w}^*$) increase their mutation rate, then the population mean fitness will increase -- a result generalized by our \emph{mean fitness principle} in Eqs.~\ref{mfp_sign} and~\ref{mfp_sign_threshold}.

Our analysis focuses on populations at equilibrium.
Nevertheless, it has been demonstrated that a modifier which increases the mutation rate of maladapted individuals can be favored by selection~\citep{Ram2012} and reduce the adaptation rate~\citep{Ram2014},
and empirical evidence suggests that \emph{stress-induced mutagenesis} is common in bacteria and yeast, and may be prevalent in plants, flies, and human cancer cells~\citep{Rosenberg2012,Fitzgerald2017b}.
Similar theoretical results have been demonstrated for a modifier that increases the recombination rate in maladapted individuals~\citep{Hadany2003a,Hadany2003b}.

>>>>

% Yoav & Lee commands
\newcommand*{\tr}{^\intercal}
\let\vec\mathbf
\newcommand{\matrx}[1]{{\left[ \stackrel{}{#1}\right]}}
\newcommand{\diag}[1]{\mbox{diag}\matrx{#1}}
\newcommand{\goesto}{\rightarrow}
\newcommand{\dspfrac}[2]{\frac{\displaystyle #1}{\displaystyle #2} }
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}
\newtheorem{lemma}{Lemma}
\newtheorem{remark}{Remark}
%\newtheorem*{mfp}{Theorem: Mean Fitness Principle}
%\newtheorem*{rvp}{Corollary: Reproductive Value Principle}
%\newtheorem*{frv}{Fisher's reproductive values}
\renewcommand\qedsymbol{} % no square at end of proof
\newcommand{\cl}{\mathbf{L}}
\newcommand{\cj}{\mathbf{J}}
\newcommand{\ci}{I}

@@

\paragraph*{General model.}\label{sec:general_model}
Consider a large haploid population with an arbitrary set of types $A_1, A_2, \ldots, A_n$.
The frequency and fitness of individuals of type $A_k$ are $f_k$ and $w_k$, respectively.
The probability that an individual of type $A_k$ will transition to some other type is $C_k$, and given a transition occurs, the probability that it will transition to type $A_j$ is $M_{j,k}$.
Therefore, the change in the frequencies of type $A_k$ is described by the transformation $f \to f'$: 
\begin{equation}
\bar{w} f'_k = (1-C_k) w_k f_k + \sum_{j=1}^{n}{C_j M_{k,j} w_j f_j},
\label{eq:model_sum}
\end{equation}
or in matrix form
\begin{equation}
\bar{w} f' = (\ci - C + MC)D f,
\label{eq:model}
\end{equation}
where $f=(f_1, f_2, \ldots, f_n)$ is a stochastic vector such that $\sum_{k=1}^n{f_k} = 1$;
$D$ is a diagonal matrix with diagonal elements $w_k$ such that $w_k \ne w_j$ for $k \ne j$, and we can assume $0 < w_k \le 1$, i.e., fitness values are positive and relative to the fitness of the fittest type; 
$C$ is a diagonal positive matrix with diagonal elements $C_k$;
$M$ is a primitive column-stochastic matrix: $M_{j,k} \ge 0$ for all $j,k$, $\sum_{j=1}^n {M_{j,k}} = 1$ for all $k$, and $M^n_{j,k} > 0$ for all $j,k$ for some natural $n$.
$\ci$ is the identity matrix;
and $\bar{w}$ is the normalizing factor such that $\sum_{k=1}^n{f'_k}=1$ and is equal to the population mean fitness (see~\nameref{sec:AppA}).

@@

The initial phenotype is acquired by learning the mature phenotype of a parent.
In individuals with initial phenotype $k$, the mature phenotype is the same as the initial phenotype with probability $1-C_k$, and is modified by individual learning or \emph{exploration}~\citep{Borenstein2008} with probability $C_k$.
Such individual exploratory learning, which can be considered either intentional or the result of incorrect learning, modifies initial phenotype $k$ to mature phenotype $j$ with probability $M_{k,j}$.

For example, the case of \emph{symmetric individual learning}~\citep{Borenstein2008} can be represented by
\begin{equation}
M_{k,j} = \begin{cases}
\frac{1}{2b+1}, & \abs{k-j} \le b \\
0, & otherwise
\end{cases},
\end{equation}
where the breadth of exploration is $b \ge 0$, and phenotypes $0$ and $n$ "absorb" transitions to $k<0$ and $k>n$, respectively. 
For instance, with $n=5$ and $b=1$ we have
\begin{equation}
M = \begin{bmatrix}
2/3 & 1/3 & 0 & 0 & 0 \\
1/3 & 1/3 & 1/3 & 0 & 0 \\
0 & 1/3 & 1/3 & 1/3 & 0 \\
0 & 0 & 1/3 & 1/3 & 1/3 \\
0 & 0 & 0 & 1/3 & 2/3
\end{bmatrix}.
\end{equation}

Therefore, if the frequency of individuals with mature phenotype $k$ in the current generation is $f_k$, then the frequency in the next generation $f'_k$ is
\begin{equation}
f'_k = (1-C_k) f_k \frac{w_{\abs{k-h}}}{\bar{w}} + \sum_{j=1}^{n}{C_j M_{k,j} \frac{w_{\abs{j-h}}}{\bar{w}} f_j },
\end{equation}
where $\bar{w} = \sum_{k=1}^{n}{f_k w_{\abs{k-h}}}$ is the population mean fitness.

@@

\begin{lemma}[Mean fitness principle]\label{lemma:mfp}
Let $\bar{w}^*$ be the leading eigenvalue of $(I-C+MC)D$,
and $u$ and $v$ be the corresponding left and right eigenvectors.
Then,
\begin{equation}\label{mfp}
\frac{\partial \bar{w}^*}{\partial C_k} = 
\frac{u_k v_k}{C_k} (\bar{w}^* - w_k),
\end{equation}
or in simpler terms,
\begin{equation}\label{mfp_sign}
sign\frac{\partial \bar{w}^*}{\partial C_k} = 
sign(\bar{w}^* - w_k).
\end{equation}

@@

We have given several examples of variation-generating processes for which this principle applies -- namely mutation, migration, and learning (see \emph{\nameref{sec:models}} section) -- but our model may apply to other processes as well.
For example, the \emph{reduction principle} applies to ecological models of dispersal, and~\citet{Gueijman2013} have demonstrated that even in homogeneous environments, fitness-associated dispersal increases the mean fitness of diploid populations and is favored by selection over uniform dispersal.
Similarly, if the transmission fidelity of culturally-transmitted traits depends on the type or fitness of the transmitting individual, we expect that our results will hold (see \emph{\nameref{sec:learn_model}} section).

Eq.~\ref{mfp} is a generalization of a result of~\citet[Eq.~4]{Ram2012}.
\citeauthor{Ram2012} modeled the accumulation of mutant alleles in a population evolving at the mutation-selection balance (see \emph{\nameref{sec:mut_model_2}} section).
Using Eq.~\ref{eq:Caswells_formula} in \nameref{sec:AppB} and a recursion on the ratios of the reproductive values \cite[see][eqs.~A5-6]{Ram2012}, they concluded that at the mutation-selection balance, if individuals with below-average fitness ($D_k < \bar{w}^*$) increase their mutation rate, then the population mean fitness will increase -- a result generalized by our \emph{mean fitness principle} in Eqs.~\ref{mfp_sign} and~\ref{mfp_sign_threshold}.

Our analysis focuses on populations at equilibrium.
Nevertheless, it has been demonstrated that a modifier which increases the mutation rate of maladapted individuals can be favored by selection~\citep{Ram2012} and reduce the adaptation rate~\citep{Ram2014},
and empirical evidence suggests that \emph{stress-induced mutagenesis} is common in bacteria and yeast, and may be prevalent in plants, flies, and human cancer cells~\citep{Rosenberg2012,Fitzgerald2017b}.
Similar theoretical results have been demonstrated for a modifier that increases the recombination rate in maladapted individuals~\citep{Hadany2003b,Hadany2003a}.
